---
title: "Linear Regression Models Project"
author: "caspar chen jc5067"
date: "11/22/2018"
output: word_document
---


# Introduction

```{r}
# data
df <- read.csv("salary.txt",header=T)
head(df)
# Names of variables 
wage <- df$wage
edu <- df$edu
exp <- df$exp
city <- df$city
reg <- df$reg
race <- df$race
deg <- df$deg
com <- df$com
emp <- df$emp

# Take a peak at the data (try other common functions) 
head(df)
dim(df)
summary(df)
```

```{r}
# Construct training and test datasets
# Use sample 
# Set seed to 0 
set.seed(0)
round(.2*nrow(df))
index <- sample(1:nrow(df),4965,replace = F)
train.data <- df[-index,]
data <- train.data 
test.data <- df[index,]

# Quality control check
sum(df$race=="black")/nrow(df)
sum(train.data$race=="black")/nrow(data)
sum(test.data$race=="black")/nrow(test.data)
table(df$race,df$city)/nrow(df)
table(train.data$race,train.data$city)/nrow(train.data)
```

# Basic EDA (exploratory data analysis)
```{r}
qqnorm(log(df$wage),ylab="log(Wage)")
boxplot(log(df$wage)~df$race,xlab="Race",ylab="log(Wage)")
summary(lm(log(df$wage)~df$race))
raceblack <- ifelse(df$race=="black","black","other")
boxplot(log(df$wage)~raceblack,xlab="Race",ylab="log(Wage)")
summary(lm(log(df$wage)~raceblack))


```

# Rough models

```{r}
# raw model 1
r.model.1 <- lm(wage~edu+exp+city+reg+race+deg+com+emp,data=train.data) 
summary(r.model.1)
AIC(r.model.1)
qqnorm(rstudent(r.model.1))
qqline(rstudent(r.model.1))
# Rough model 2 
r.model.2 <- lm(log(wage)~edu+exp+city+reg+race+deg+com+emp,data=train.data)
summary(r.model.2)
AIC(r.model.2)
qqnorm(rstudent(r.model.2),ylab="log(Wage)")
qqline(rstudent(r.model.2))
```

#Test log(wage) with every x variable
```{r}
par(mfrow = c(2,4))

# edu
plot((data$edu),log(data$wage),xlab="Edu",ylab="log(Wage)")
lines(supsmu((data$edu),log(data$wage)),col=2)

# exp
plot(data$exp,log(data$wage),xlab="Exp",ylab="log(Wage)")
lines(supsmu(data$exp,log(data$wage)),col=2)
 
#city
boxplot(log(data$wage)~data$city,xlab="City",ylab="log(Wage)")
#reg
boxplot(log(data$wage)~data$reg,xlab="Reg",ylab="log(Wage)")

#race
boxplot(log(data$wage)~data$race,xlab="Race",ylab="log(Wage)")

#deg
boxplot(log(data$wage)~data$deg,xlab="Deg",ylab="log(Wage)")

# com
plot(data$com,log(data$wage),xlab="Com",ylab="log(Wage)")
lines(supsmu(data$com,log(data$wage)),col=2)
# emp
plot(data$emp,log(data$wage),xlab="Emp",ylab="log(Wage)")
lines(supsmu(data$emp,log(data$wage)),col=2)

```

```{r}
# Rough model 3 
r.model.3 <- lm(log(wage)~edu+poly(exp, degree = 2)+city+reg+race+deg+emp,data=train.data)
summary(r.model.3)
AIC(r.model.3)
qqnorm(rstudent(r.model.3),ylab="log(Wage)")
qqline(rstudent(r.model.3))
```
# Look for interactions using plots and AIC 

```{r}
#Categorical varaibles vs Categorical varaibles
# City vs region
# There exists an interaction and the level of region that interacts with city is west.
interaction.plot(city,reg,log(wage), main="Interaction between region and city")
summary(lm(log(data$wage) ~ city  + reg + city*reg, data = data))

# City vs degree
# There exists an interaction
interaction.plot(city,deg,log(wage), main="Interaction between degree and city")
summary(lm(log(data$wage) ~city + deg + city*deg , data = data))

# region vs degree
#There exists an interaction
reg <- data$reg
deg <- data$deg
wage <- data$wage
interaction.plot(reg,deg,log(wage))
```

```{r}
#Categorical varaibles vs Continuous varaibles
# race vs edu
plot(data$edu,log(wage),col=data$race)
plot(data$edu,log(wage),col="lightgrey")
black <- data$race=="black"
white <- data$race=="white"
other <- data$race=="other"
# smoother
plot(data$edu,log(wage),col="lightgrey")
abline(lm(log(df$wage)[black]~df$edu[black]),col=2)
abline(lm(log(df$wage)[white]~df$edu[white]),col=3)
abline(lm(log(df$wage)[other]~df$edu[other]),col=4)
legend("topleft",legend=c("Black","White","Other"),fill=2:(length(levels(df$race))+1))
# Looks close to parallel. An interaction between race and education probably shouldn’t be included.

# City vs edu

plot(data$edu,log(data$wage),col=data$city)
plot(data$edu,log(data$wage),col="lightgrey")
yes <- data$city=="yes"
no <- data$city=="no"
# smoother
plot(data$edu,log(data$wage),col="lightgrey",xlab="Edu")
lines(supsmu(data$edu[yes],log(data$wage)[yes]),col=2)
lines(supsmu(data$edu[no],log(data$wage)[no]),col=3)
legend("topright",legend=c("Yes","No"),col=c(2,3),lty=c(1,1))

# City vs exp
plot(data$exp,log(data$wage),col=data$city)
plot(data$exp,log(data$wage),col="lightgrey")
yes <- data$city=="yes"
no <- data$city=="no"
# smoother
plot(data$exp,log(data$wage),col="lightgrey",xlab="Exp")
lines(supsmu(data$exp[yes],log(data$wage)[yes]),col=2)
lines(supsmu(data$exp[no],log(data$wage)[no]),col=3)
legend("topright",legend=c("Yes","No"),col=c(2,3),lty=c(1,1))

# Reg vs edu
plot(data$edu,log(data$wage),col=data$reg)
plot(data$edu,log(data$wage),col="lightgrey")

midwest <- data$reg=="midwest"
northeast <- data$reg=="northeast"
south <- data$reg=="south"
west <- data$reg=="west"

# smoother
plot(data$edu,log(data$wage),col="lightgrey",xlab="Edu")
lines(supsmu(data$edu[midwest],log(data$wage)[midwest]),col=2)
lines(supsmu(data$edu[northeast],log(data$wage)[northeast]),col=3)
lines(supsmu(data$edu[south],log(data$wage)[south]),col=4)
lines(supsmu(data$edu[west],log(data$wage)[west]),col=5)
legend("topright",legend=c("Midwest","Northeast","South","West"),col=c(2,3,4,5),lty=c(1,1,1,1))

# Reg vs exp
plot(data$exp,log(data$wage),col=data$reg)
plot(data$exp,log(data$wage),col="lightgrey")
midwest <- data$reg=="midwest"
northeast <- data$reg=="northeast"
south <- data$reg=="south"
west <- data$reg=="west"
# smoother
plot(data$exp,log(data$wage),col="lightgrey",xlab="Exp")
lines(supsmu(data$exp[midwest],log(data$wage)[midwest]),col=2)
lines(supsmu(data$exp[northeast],log(data$wage)[northeast]),col=3)
lines(supsmu(data$exp[south],log(data$wage)[south]),col=4)
lines(supsmu(data$exp[west],log(data$wage)[west]),col=5)
legend("topright",legend=c("Midwest","Northeast","South","West"),col=c(2,3,4,5),lty=c(1,1,1,1))

# Deg vs edu
plot(data$edu,log(data$wage),col=data$deg)
plot(data$edu,log(data$wage),col="lightgrey")
yes <- data$deg=="yes"
no <- data$deg=="no"
# smoother
plot(data$edu,log(data$wage),col="lightgrey",xlab="Edu")
lines(supsmu(data$edu[yes],log(data$wage)[yes]),col=2)
lines(supsmu(data$edu[no],log(data$wage)[no]),col=3)
legend("topright",legend=c("Yes","No"),col=c(2,3),lty=c(1,1))

```

```{r}
# Pearson correlation for continuous variable only
cor(data.frame(log(data$wage), data$edu,data$exp, data$emp))
```


```{r}
#final model
final.model <- lm(log(wage)~ edu + poly(exp,2) + city + reg + race + deg + emp + edu*exp + edu*city + edu*deg + exp*deg + city*reg + city*deg + reg*deg, data=train.data)
summary(final.model)
AIC(final.model)
#research question 2 
train.data2 <- train.data
train.data2$race <- ifelse(train.data$race=="black","black","other")
final.model2 <- lm(log(wage)~ edu + poly(exp,2) + city+reg + race + deg + emp + edu*exp + edu*city + edu*deg + exp*deg + city*reg + city*deg + reg*deg, data=train.data2)
summary(final.model2)
AIC(final.model2)
AIC(final.model2) < AIC(final.model)
```

# Model Validation
## i.Diagnostics
```{r}
par(mfrow=c(2,3))
#1.Scatterplot
plot(wage~race,data=train.data,main="Scatter-Plot")
abline(final.model,col =2)
#2.QQplot
qqnorm(rstudent(final.model),main="QQ-Plot")
abline(a=0,b=1,lty=1,col=2)
#3.histogram
hist(rstudent(final.model), main ="Histogram", xlab ="Studentized Deleted Residuals")
#4.line plot of the studentized deleted residuals
plot(rstudent(final.model),main="Line Plot",ylab="Studentized Deleted Residuals")
abline(h=0,lty=1,col =2)
lines(1:19858,rstudent(final.model),col=2)
#5.Studentized deleted residuals verses predicted values y_hat
plot(predict(final.model),rstudent(final.model),main="Residual Plot",xlab="Y-hat",ylab="Studentized Deleted Residuals")
abline(h=0,lty=2)
lines(supsmu(predict(final.model),rstudent(final.model)),col=2)
#6.Studentized deleted residuals verses covariate race
plot(rstudent(final.model)~train.data$race,main="Residual Plot",xlab="race",ylab="Studentized Deleted Residuals")
abline(h=0,lty=1,col =2)
```


## ii.Compute MSPR and MSE

Below we compute the MSPR using our final model trained from the training set on the test set. First fit the final model on the training set. 
```{r}
# Compute MSE 
MSE <- sum((residuals(final.model))^2)/(nrow(train.data)-16)

final.modelfulldata <- lm(log(wage)~ edu + poly(exp,2) + city + reg + race + deg + emp + edu*exp + edu*city + edu*deg + exp*deg + city*reg + city*deg + reg*deg, data=df)
# For comparison, we can compute MSE of the earlier final model
MSE.earler <- sum((residuals(final.modelfulldata))^2)/(nrow(df)-16)
```


Next we have to extract the test data. Then plug the test data in the predict function to find the Y-predictions, i.e., Y.test. Then construct the MSPR and compare to MSE.   

```{r}

# Note we have to take out the wage variable, column 1 of test data
names(train.data[,-1])

Y.test <- test.data[,1]
X.test <- test.data[,-1]
n.test <- nrow(X.test)
n.test
Y.hat.test <- predict(final.model, newdata = X.test)
length(Y.hat.test)==n.test
# MSPR
MSPR <- mean((log(Y.test)-Y.hat.test)^2)
MSPR
MSE
MSE.earler
round(c(MSPR=MSPR,MSE=MSE,MSEearler=MSE.earler),4)
```

## iii.influential Observations

```{r}
#DFBETAS
final.model$coef
#DFBETAS-raceother
dfbetas(final.model)[,9]
n <- length(train.data$wage)
plot(dfbetas(final.model)[,9],main="DFBETAS-Raceother",ylim=c(-.4,.4))
abline(h=2/sqrt(n),col = 2)
abline(h=-2/sqrt(n),col = 2)
#DFBETAS-racewhite
dfbetas(final.model)[,10]
plot(dfbetas(final.model)[,10],main="DFBETAS-Racewhite",ylim=c(-.4,.4))
abline(h=2/sqrt(n),col = 2)
abline(h=-2/sqrt(n),col = 2)
```
 
## iv. VIF
```{r}
#Variance Inflation Factor (VIF)
library(faraway) 
df$reg_num <- as.numeric(factor(df$reg, levels=c("northeast" ,"midwest", "south", "west")))
df$city_num <- as.numeric(factor(df$city, levels=c("yes" , "no")))
df$deg_num <- as.numeric(factor(df$deg, levels=c("yes" ,"no")))
df$race_num <- as.numeric(factor(df$race, levels=c("black" ,"white", "other")))

df.vif <- data.frame(edu,poly(exp,2), emp,df$reg_num,df$city_num,df$deg_num,df$race_num,edu*exp, edu*df$city_num, edu*df$deg_num, exp*df$deg_num, df$city_num*df$reg_num, df$city_num*df$deg_num, df$reg_num*df$deg_num)
vif(df.vif)
```



